# Case study: range size and climatic niche breadth of European plants {#sec:case}

## Question and data

Species with broad environmental tolerances are expected to occupy larger
geographical ranges. This is the niche-breadth hypothesis for range size
[@brown1984abundance], and a meta-analysis found the relationship across
most taxa and regions tested [@slatyer2013niche]. Testing it on
occurrence data requires, for every species, its range size and the breadth
of climate it occupies. Both are computed from all of the species' records,
which for European vascular plants means several hundred million rows. Both
also increase with the number of records, because a species recorded more
often is found in more cells and in more extreme climates. Sampling effort
is well known to be uneven across GBIF data [@beck2014bias; @meyer2015gbif],
so it has to enter the model.

The records come from the GBIF occurrence snapshot of 1 September 2026
[@gbif2026snapshot], which GBIF distributes as Parquet files on public cloud
storage. We selected the records of Tracheophyta whose country code is
European, `r fmt(CS$n_records / 1e6, 1)` million rows, and kept
`r CS$n_cols` columns: species key and name, coordinates, coordinate
uncertainty, year, basis of record, occurrence status, taxon rank, country,
and dataset key. Climate comes from WorldClim 2.1 [@fick2017worldclim] at 10
arc-minutes: annual mean temperature (BIO1) and annual precipitation
(BIO12). Country and region come from the Natural Earth 1:10m country
polygons [@naturalearth], grouped into the four European subregions of the
United Nations geoscheme that Natural Earth records.

## Workflow

The analysis has six stages, each a separate R script run in a fresh
process (Table \@ref(tab:stages-tab)).

0. **Acquisition** (outside vectra). A DuckDB query over the snapshot's
   Parquet parts selected the rows and columns above and wrote them to
   `r CS$n_raw_files` local Parquet files. In practice this step is a GBIF
   download request, for example through \CRANpkg{rgbif}
   [@chamberlain2025rgbif].
1. **Ingest.** vectra does not read Parquet, so each file was read with
   arrow and appended to a single `.vtr` store with `append_vtr()`.
2. **Clean.** Records were kept if present, identified to species, collected
   in or after 1950, with a coordinate uncertainty of at most 10 km or none
   given, of an observation or specimen basis, and inside the climate grid.
   Each record was assigned the key of the 10-arc-minute cell containing it,
   computed from its coordinates in `mutate()`. The result streams to disk.
3. **Cells.** The two climate layers were read with `tbl_tiff()` as tables of
   pixels, given the same cell key, and joined. The cell centres were tagged
   with country and region by `spatial_join()` against the country polygons,
   dropping cells outside Europe. Each cell's area was computed from its
   latitude.
4. **Occupancy.** The cleaned records were reduced to one row per species and
   cell (the number of records, the number of datasets, and the first and
   last year), then joined to the cell table.
5. **Models.** The occupancy table was partitioned on disk by region with
   `offload(by = "region")`. `group_modify()` then read each region's shard
   into R, reduced it to one row per species, and fitted a linear model. A
   Europe-wide species table was computed in the engine with a grouped
   `summarise()` and modelled the same way.

For each species we computed range size as the summed area of occupied cells
(km²). Niche breadth for temperature and for log precipitation is the
standard deviation of the occupied cells' values. Within regions it was
rarefied to 10 cells (the mean over 20 random draws of 10 occupied cells),
so that breadth is measured at the same sample size for every species. The
Europe-wide table uses the unrarefied standard deviation, which the engine
computes in a grouped `summarise()`. We then fitted

$$\log(\text{range}) = \beta_0 + \beta_T \log(\text{breadth}_T) +
\beta_P \log(\text{breadth}_P) + \beta_n \log(n_\text{records}) + \varepsilon$$

to species with at least 50 records in at least 10 cells. The coefficients
$\beta_T$ and $\beta_P$ are elasticities: the proportional change in range
size for a proportional change in niche breadth, at a given number of
records.

```{r stages-tab}
st <- CS$stages
if (!is.null(st)) {
  knitr::kable(st, caption = "Resource use of the case-study stages. Each stage ran in a fresh R process on the 2-vCPU, 8-GB machine; peak memory is the maximum resident set size of the process; disk is the size of the stage's output.",
               booktabs = TRUE, digits = 1)
}
```

## Computational cost

\textbf{TODO (pending run).} Stage 0 finished for 44 of the 122 source
groups before the session stopped. Stages 1 to 5 have been tested
end-to-end on a 0.87-million-record sample but not yet run on the full
extract. Fill Table \@ref(tab:stages-tab) from `casestudy/stages.jsonl` and
report the total time, peak memory per stage, and disk use.

## Result

\textbf{TODO (pending run).} Report $\beta_T$ and $\beta_P$ by region with
95% intervals (the figure below), the number of species per region, and the
comparison with the unrarefied Europe-wide model. Relate the sign and size
of the elasticities to @slatyer2013niche.

```{r fits-fig, fig.cap = CS$fig_caption, fig.width = 7, fig.height = 3.2, eval = !is.null(CS$fits)}
source("scripts/fig_case.R")
```
