# Comparative evaluation {#sec:eval}

```{r eval-setup, include = FALSE}
bd <- readRDS(file.path(.dir, "bench_data.rds"))
nrep <- max(runs$rep)
S <- function(tool, w, p, stat = "wall") B(tool, w, p, stat)
R <- function(tool, w, p) S(tool, w, p, "rss") / 1024        # GB
```

## Claims and experiments

Each claim of Sections \@ref(sec:design) and \@ref(sec:memory) is paired with
an experiment in Table \@ref(tab:claims-tab). All experiments were run on one
machine with two virtual CPUs and 8 GB of memory without swap (Section
\@ref(sec:details)). A table larger than about 6 GB cannot be held in R there
at all, and every tool must work within the same limit.

```{r claims-tab}
claims <- data.frame(
  Claim = c("Streaming operations use memory independent of input size",
            "Stateful operations are bounded by the budget, not by groups or build-side rows",
            "Collecting a result costs memory in proportion to its size; sinks do not",
            "Pushdown, zone maps, pruning, and indexes reduce the data read",
            "Offloading a prepared query reduces the cost of iterative fits",
            "Streaming spatial joins remain practical beyond in-memory sizes",
            "The components compose into a complete analysis"),
  Experiment = c("scan, filter, and aggregate 10 to 100 million rows",
                 "grouping into 1,000 to 10 million groups; joins with 1,000 to 50 million build rows; full sort",
                 "keep 0.1% to 50% of 50 million rows, then collect() or write to a file",
                 "paired queries on one 50-million-row store (vectra only)",
                 "logistic regression by IRLS over a joined, derived table",
                 "point-in-polygon tagging of 1 to 50 million points",
                 "401 million GBIF records to per-region models"),
  Evidence = c("Input size and streaming", "Stateful operations", "Input size and streaming",
               "Optimiser ablation", "Repeated passes for model fitting", "Spatial joins",
               "Case study"),
  check.names = FALSE)
ktab(claims, "Claims and the experiments that test them, with the subsection that reports each.", widths = c("5.4cm", "5.4cm", "3.2cm"))
```

## Setup

**Data.** The tabular benchmarks use a synthetic table shaped like an
occurrence dataset. It has 11 columns: a row identifier, a year, four
integer keys with 10^3^, 10^5^, 10^6^, and 10^7^ distinct values, a species
name drawn from 10^4^ strings, two numeric measurements, and two
coordinates. The table has 10^8^ rows, written in ten files of 10^7^ rows in
each tool's own prepared format. That is `.vtr` for vectra
(`r fmt(bd$sizes_gb[["vtr"]], 1)` GB in total), Parquet with arrow's default
compression for arrow and DuckDB (`r fmt(bd$sizes_gb[["pq"]], 1)` GB), and
\CRANpkg{fst} [@fst] for data.table (`r fmt(bd$sizes_gb[["fst"]], 1)` GB). An
input of N rows is the first N/10^7^ files. vectra combines them with
`bind_rows()`, arrow with `open_dataset()`, DuckDB with a list passed to
`read_parquet()`, and data.table reads them with `read_fst()` and binds them
with `rbindlist()`, reading only the columns a query uses. The spatial
benchmark tags the table's coordinates, which are uniform over a box from
10°W to 30°E and 36°N to 70°N, with 53 country polygons from Natural Earth
1:10m clipped to the same box.

**Queries.** Each workload was written the way a user of each package would
write it: dplyr verbs for vectra and arrow, SQL for DuckDB, and `[.data.table`
syntax for data.table. No query was rewritten to suit one engine. Results
that are large (grouped results, sorted tables) were written to each tool's
own file format, and small results were collected into R. For every
configuration, the tools that completed returned the same checksum of their
result (row counts and rounded sums).

**Measurement.** Every run was a fresh `Rscript` process started after the
operating-system page cache had been dropped, so all reads were cold. Every
tool was limited to two threads. Memory settings were left at each package's
defaults. For vectra that is a budget of half the physical memory
(`r fmt(4211091456 / 1024^3, 1)` GB). A second configuration, "vectra (1 GB
budget)", sets `options(vectra.memory = "1GB")`. Elapsed time and peak
resident memory were recorded by GNU `time`, which reports the maximum
resident set size of the process and therefore includes every library's
native allocations. Temporary disk was the largest size of the process's
temporary directory, sampled every 0.5 seconds. Bytes read were taken from
`/proc/self/io`. Each configuration was repeated `r nrep` times, and figures
show the median. A configuration whose first run was killed for lack of
memory was not repeated, and the two slowest spatial runs were run once.
Runs killed by the kernel's out-of-memory handler are shown as crosses at
the memory they had reached.

## Input size and streaming

```{r fig-scan, fig.cap = "Scan, filter, derive, and aggregate by year (75 groups) as input grows from 10 to 100 million rows. (A) Elapsed time. (B) Peak resident memory of the R process. The dashed line is the machine's physical memory.", out.width = "100%"}
knitr::include_graphics("figures/scan.pdf")
```

The first workload filters on two columns, derives a product, and aggregates
by year, returning 75 rows (Figure \@ref(fig:fig-scan)). DuckDB was fastest
at every size (`r fmt(S("duckdb", "scan", "1e8"))` s for 10^8^ rows) and used
the least memory (`r fmt(R("duckdb", "scan", "1e8"), 2)` GB). arrow and
data.table followed in time. data.table's memory grew linearly with input,
since it loads the three columns it needs, and reached
`r fmt(R("datatable", "scan", "1e8"))` GB at 10^8^ rows. vectra took
`r fmt(S("vectra", "scan", "1e8"))` s at 10^8^ rows. Its memory also grew
with input, to `r fmt(R("vectra", "scan", "1e8"))` GB, above its nominal
budget. This is not a streaming cost. The grouped `summarise()` sorts its
input, and because the filter is followed by a `mutate()`, column pruning
stopped there and the sort carried all 12 columns of every surviving row
(Section \@ref(sec:design)). With a 1 GB budget, vectra's memory stayed at
`r fmt(R("vectra1g", "scan", "5e7"), 1)` to `r fmt(R("vectra1g", "scan", "1e8"), 1)` GB
from 5 × 10^7^ rows upwards. It spilled `r fmt(S("vectra1g", "scan", "1e8", "tmp") / 1024, 1)` GB
to disk at 10^8^ rows, and time rose to `r fmt(S("vectra1g", "scan", "1e8"))` s.

```{r fig-collect, fig.cap = "Memory needed to materialise a result in R. A filter keeps a fraction of 50 million rows and four columns; the result is either collected into R (solid) or written to a file (dashed).", out.width = "70%"}
knitr::include_graphics("figures/collect.pdf")
```

Pure streaming shows in Figure \@ref(fig:fig-collect). A filter keeps
between 0.1% and 50% of 5 × 10^7^ rows and four columns. When the result was
written to a file, vectra's peak memory stayed between
`r fmt(min(summ$rss[summ$tool == "vectra" & summ$workload == "sink"]), 0)` and
`r fmt(max(summ$rss[summ$tool == "vectra" & summ$workload == "sink"]), 0)` MB
at every selectivity, the least of the three tools. DuckDB was similar
(`r fmt(max(summ$rss[summ$tool == "duckdb" & summ$workload == "sink"]), 0)` MB
at most), and arrow used `r gb(max(summ$rss[summ$tool == "arrow" & summ$workload == "sink"]))` GB.
Collecting the result costs memory in proportion to its size for every tool.
vectra needed `r gb(S("vectra", "collect", "0.5", "rss"))` GB to return
25 million rows, arrow `r gb(S("arrow", "collect", "0.5", "rss"))` GB, and
DuckDB `r gb(S("duckdb", "collect", "0.5", "rss"))` GB. Writing the half-size
result with vectra's default `compress = "fast"` was slower than collecting
it (`r fmt(S("vectra", "sink", "0.5"))` s against
`r fmt(S("vectra", "collect", "0.5"))` s), because the encoder compresses
every block.

## Stateful operations

```{r fig-stateful, fig.cap = "Grouping and joining 50 million rows. (A, B) Group by a key with 1,000 to 10 million distinct values and aggregate three columns; the result is written to a file. (C, D) Inner join to a build side of 1,000 to 50 million rows, then aggregate by year. Crosses mark runs killed for lack of memory.", out.width = "100%"}
knitr::include_graphics("figures/stateful.pdf")
```

Grouping and joining (Figure \@ref(fig:fig-stateful)) show the consequences
of the sort-based design most clearly. vectra's memory for grouping did not
depend on the number of groups (panel B). At the default budget it was
`r gb(S("vectra", "groups", "1e3", "rss"))` GB for 10^3^ groups and
`r gb(S("vectra", "groups", "1e7", "rss"))` GB for 10^7^. With a 1 GB budget
it was `r gb(S("vectra1g", "groups", "1e7", "rss"))` GB at every group
count, with `r fmt(S("vectra1g", "groups", "1e7", "tmp") / 1024, 1)` GB spilled.
The hash-based engines grew with the number of groups: arrow from
`r gb(S("arrow", "groups", "1e3", "rss"))` to `r gb(S("arrow", "groups", "1e7", "rss"))` GB,
and DuckDB from `r fmt(S("duckdb", "groups", "1e3", "rss"), 0)` MB to
`r gb(S("duckdb", "groups", "1e7", "rss"))` GB. Even at 10^7^ groups,
however, both stayed below vectra's default peak. The price of the sort is
time. vectra took `r fmt(S("vectra", "groups", "1e3"))` s to aggregate into
10^3^ groups, where DuckDB took `r fmt(S("duckdb", "groups", "1e3"))` s and
arrow `r fmt(S("arrow", "groups", "1e3"))` s. With a 1 GB budget, the extra
read and write of the spill added about 25 to 30 s at every group count.

For joins, all three engines keep the build side in memory while it fits,
and memory rose with its size (panel D). At 5 × 10^7^ build rows, as many as
the probe side, vectra at its default budget was killed for lack of memory at
`r gb(S("vectra", "join", "5e7", "rss"))` GB. With a 1 GB budget it
completed by partitioning both sides to disk, but its peak of
`r gb(S("vectra1g", "join", "5e7", "rss"))` GB was five times the budget.
The Grace partitioning bounds the hash table, but not the sort for the
grouped aggregation that follows it. arrow was at the limit of the machine:
it completed one of its three runs, at `r gb(S("arrow", "join", "5e7", "rss"))` GB,
and was killed in the other two. DuckDB completed at
`r gb(S("duckdb", "join", "5e7", "rss"))` GB, and data.table at
`r gb(S("datatable", "join", "5e7", "rss"))` GB.

```{r sort-tab}
st <- summ[summ$workload == "sort", ]
st$tool <- factor(st$tool, levels = names(tool_lab))
st <- st[order(st$param_num, st$tool), ]
tab <- data.frame(Rows = paste0(st$param_num / 1e6, "M"), Tool = tool_lab[as.character(st$tool)],
                  `Time (s)` = ifelse(st$n_ok > 0, sprintf("%.1f", st$wall), "killed (memory)"),
                  `Peak memory (GB)` = sprintf("%.2f", st$rss / 1024),
                  `Temporary disk (GB)` = ifelse(st$n_ok > 0, sprintf("%.2f", st$tmp / 1024), ""),
                  check.names = FALSE)
ktab(tab, "Sorting all 11 columns by a numeric column and writing the result to a file.")
```

Sorting (Table \@ref(tab:sort-tab)) is the operation where external memory
is unavoidable once the table exceeds memory. At 10^7^ rows, data.table and
DuckDB sorted in about 5 s. vectra was the slowest tool at this size, and its
in-memory path (`r fmt(S("vectra", "sort", "1e7"))` s) was slower than
its spilling path with a 1 GB budget (`r fmt(S("vectra1g", "sort", "1e7"))` s).
At 5 × 10^7^ rows, which is about 4 GB of data in R, arrow and data.table
were killed for lack of memory. DuckDB completed in
`r fmt(S("duckdb", "sort", "5e7"))` s with a peak of
`r gb(S("duckdb", "sort", "5e7", "rss"))` GB. vectra took
`r fmt(S("vectra", "sort", "5e7"))` s at either budget, and with a 1 GB budget
its peak stayed at `r gb(S("vectra1g", "sort", "5e7", "rss"))` GB. At 10^8^
rows the two external sorts differed mainly in their disk use. DuckDB
completed in `r fmt(S("duckdb", "sort", "1e8"))` s, but it needed
`r fmt(S("duckdb", "sort", "1e8", "tmp") / 1024, 1)` GB of temporary files in
addition to the `r fmt(S("duckdb", "sort", "1e8", "out") / 1024, 1)`-GB
output. Its first two attempts failed because the 9 GB of free disk space
then available was not enough, and the reported run was made after freeing
space. vectra needed `r fmt(S("vectra", "sort", "1e8", "tmp") / 1024, 1)` GB
of temporary files. It completed in `r fmt(S("vectra1g", "sort", "1e8"))` s
with a peak of `r gb(S("vectra1g", "sort", "1e8", "rss"))` GB under the 1 GB
budget, and `r gb(S("vectra", "sort", "1e8", "rss"))` GB under the default.

## Optimiser ablation

```{r abl-tab}
ab <- function(p) c(S("vectra", "abl", p), S("vectra", "abl", p, "rchar"))
rows <- list(
  c("Zone maps", "filter on a column ordered across row groups", "zm_sorted"),
  c("", "same selectivity (1%), column in random order", "zm_random"),
  c("Column pruning", "sum of 1 column", "cols_1"),
  c("", "sums of 10 columns", "cols_all"),
  c("Hash index", "point lookup with a .vtri index", "idx_vtri"),
  c("", "same lookup without an index", "idx_none"),
  c("Pushdown order", "filter() before mutate()", "mut_after"),
  c("", "the same filter after mutate()", "mut_before"))
abt <- do.call(rbind, lapply(rows, function(r) {
  v <- ab(r[3]); data.frame(Mechanism = r[1], Query = r[2],
    `Time (s)` = sprintf("%.2f", v[1]), `Data read (MB)` = sprintf("%.0f", v[2]), check.names = FALSE)
}))
ktab(abt, "Paired queries on a single 50-million-row .vtr store (381 row groups). Each pair returns the same result. Data read is the number of bytes the process read from files.")
```

Table \@ref(tab:abl-tab) isolates the optimiser's mechanisms on a single
store of 5 × 10^7^ rows. Zone maps skipped almost all row groups for a
filter on a column whose values are ordered across row groups. The query
read `r fmt(S("vectra", "abl", "zm_sorted", "rchar"), 0)` MB and took
`r fmt(S("vectra", "abl", "zm_sorted"), 2)` s. The same selectivity on a
randomly ordered column read every row group of the two columns involved
(`r fmt(S("vectra", "abl", "zm_random", "rchar"), 0)` MB). Column pruning
scaled the data read with the number of columns used. A `.vtri` index
reduced a point lookup from `r fmt(S("vectra", "abl", "idx_none"), 1)` s and
`r gb(S("vectra", "abl", "idx_none", "rchar"))` GB read to
`r fmt(S("vectra", "abl", "idx_vtri"), 2)` s and
`r fmt(S("vectra", "abl", "idx_vtri", "rchar"), 0)` MB. The last pair shows
the cost of the optimiser's locality. A filter written after a `mutate()`
is not pushed into the scan, and the `mutate()` blocks pruning. The query
then read the whole store and took
`r ratio(S("vectra", "abl", "mut_before"), S("vectra", "abl", "mut_after"), 0)`
times as long as the same query with the verbs in the other order.

## Repeated passes for model fitting

```{r model-tab}
mt <- summ[summ$workload %in% c("model_rebuild", "model_offload", "model_glm"), ]
mt$Approach <- c(model_glm = "collect(), then glm()", model_rebuild = "bigglm(), plan rebuilt each pass",
                 model_offload = "offload() once, then bigglm()")[mt$workload]
mt <- mt[order(mt$param_num, match(mt$workload, c("model_glm", "model_rebuild", "model_offload"))), ]
tab <- data.frame(Rows = paste0(mt$param_num / 1e6, "M"), Approach = mt$Approach,
                  `Time (s)` = ifelse(mt$n_ok > 0, sprintf("%.1f", mt$wall), "killed (memory)"),
                  `of which preparation (s)` = ifelse(mt$n_ok > 0 & mt$workload != "model_rebuild", sprintf("%.1f", mt$t_prep), ""),
                  `Passes` = ifelse(mt$n_ok > 0, sprintf("%.0f", mt$iter), ""),
                  `Peak memory (GB)` = sprintf("%.2f", mt$rss / 1024),
                  `Data read (GB)` = ifelse(mt$n_ok > 0, sprintf("%.1f", mt$rchar / 1024), ""),
                  check.names = FALSE)
ktab(tab, "Logistic regression on a prepared query (filter, join to a table of one million rows, derived response and predictors). Rows are fact-table rows before filtering. Passes are IRLS iterations to convergence.")
```

A generalised linear model fitted by iteratively reweighted least squares
reads its data once per iteration. Table \@ref(tab:model-tab) fits a
logistic regression to a query that filters the fact table, joins it to a
table of 10^6^ rows, and derives the response. Fed to `biglm::bigglm()`
through a plan factory, every iteration re-ran the scan and the join. For
5 × 10^7^ fact rows the fit took `r fmt(S("vectra", "model_rebuild", "5e7"))` s
and read `r gb(S("vectra", "model_rebuild", "5e7", "rchar"))` GB. Offloading
the prepared query once took `r fmt(S("vectra", "model_offload", "5e7", "t_prep"))` s,
after which each pass read the three prepared columns. The whole fit then
took `r fmt(S("vectra", "model_offload", "5e7"))` s at
`r fmt(S("vectra", "model_offload", "5e7", "rss"), 0)` MB. Collecting the
prepared data and calling `glm()` was a reasonable alternative at 10^7^ rows,
though it needed `r gb(S("vectra", "model_glm", "1e7", "rss"))` GB. At
5 × 10^7^ rows it was killed for lack of memory. All three approaches gave
the same coefficients to ten decimal places.

## Spatial joins

```{r fig-pip, fig.cap = "Tagging points with the European country that contains them (53 polygons) and counting points per country. vectra streams the points against GEOS prepared geometries; sf builds one sf object of all points and joins it in memory; DuckDB uses its spatial extension's spatial join. Crosses mark runs killed for lack of memory.", out.width = "100%"}
knitr::include_graphics("figures/pip.pdf")
```

The spatial benchmark (Figure \@ref(fig:fig-pip)) tags each point with the
country polygon containing it and counts points per country. At 10^6^
points all three tools completed, vectra in `r fmt(S("vectra", "pip", "1e6"))` s,
\pkg{sf} in `r fmt(S("sf", "pip", "1e6"))` s, and DuckDB in
`r fmt(S("duckdb", "pip", "1e6"))` s. From 10^7^ points upwards, building
the \pkg{sf} point object and its join index exceeded memory, and the process
was killed at about 6 GB. vectra's time grew linearly in the number of
points, at about `r fmt(5e7 / S("vectra", "pip", "5e7") / 1e3, 0)` thousand
points per second on two threads, and it tagged 5 × 10^7^ points in
`r fmt(S("vectra", "pip", "5e7") / 60, 1)` minutes. Its peak memory rose from
`r gb(S("vectra", "pip", "1e7", "rss"))` to `r gb(S("vectra", "pip", "5e7", "rss"))` GB.
That rise comes from the output buffer, which holds joined rows as R data
frames until they reach an eighth of the budget, and not from the input
size. The spilled run files, which include a WKB point geometry for every
output row, took `r fmt(S("vectra", "pip", "5e7", "tmp") / 1024, 1)` GB of
temporary disk. DuckDB's spatial join used the least memory
(`r fmt(S("duckdb", "pip", "5e7", "rss"), 0)` MB) but was about three times
slower (`r fmt(S("duckdb", "pip", "5e7") / 60, 1)` minutes for 5 × 10^7^
points). Its plan indexes the polygons' bounding boxes and then tests each
candidate pair with an unprepared exact predicate. vectra prepares each
polygon once. At every size, all tools that completed matched the same number
of points and found them in the same number of countries.
