---
title: "Getting Started with vectra"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with vectra}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

vectra is a columnar query engine for R that handles larger-than-RAM datasets
with familiar dplyr-style verbs. All operations are lazy -- nothing runs until
you call `collect()`.

## Installation

Install the released version from CRAN:

```r
install.packages("vectra")
```

Or the development version from GitHub:

```r
# install.packages("devtools")
devtools::install_github("gcol33/vectra")
```

## Writing and reading data

vectra uses its own binary columnar format (`.vtr`) for fast, memory-efficient
storage. Write any data.frame with `write_vtr()` and open it lazily with `tbl()`:

```{r write-read}
library(vectra)

f <- tempfile(fileext = ".vtr")
write_vtr(mtcars, f)

# tbl() returns a lazy reference -- no data is loaded yet
node <- tbl(f)
node
```

Call `collect()` to materialize the result as a data.frame:

```{r collect}
tbl(f) |> collect() |> head()
```

## Filtering and selecting

Use `filter()` to keep rows matching a condition, and `select()` to pick columns.
Tidyselect helpers like `starts_with()` and `everything()` work inside `select()`.

```{r filter-select}
tbl(f) |>
  filter(cyl == 6, mpg > 19) |>
  select(mpg, cyl, hp, wt) |>
  collect()
```

Multiple conditions in `filter()` are combined with AND. Use `|` for OR:

```{r filter-or}
tbl(f) |>
  filter(cyl == 4 | cyl == 8) |>
  select(mpg, cyl) |>
  collect() |>
  head()
```

Use `%in%` for set membership and `if_else()` inside `mutate()` for conditional
values:

```{r filter-in}
tbl(f) |>
  filter(cyl %in% c(4, 8)) |>
  mutate(size = if_else(cyl > 6, "large", "small")) |>
  select(mpg, cyl, size) |>
  collect() |>
  head()
```

## Transforming columns

`mutate()` adds or replaces columns. Arithmetic, comparison, and boolean
operators all work:

```{r mutate}
tbl(f) |>
  mutate(kpl = mpg * 0.425144, hp_per_wt = hp / wt) |>
  select(mpg, kpl, hp, wt, hp_per_wt) |>
  collect() |>
  head()
```

`transmute()` works like `mutate()` but keeps only the new columns:

```{r transmute}
tbl(f) |>
  transmute(efficiency = mpg / wt, power_ratio = hp / disp) |>
  collect() |>
  head()
```

For string operations, write a small dataset with character columns:

```{r string-ops}
people <- data.frame(
  name = c("Alice", "Bob", "Charlie", "Diana"),
  city = c("Amsterdam", "Berlin", "Chicago", "Dublin"),
  stringsAsFactors = FALSE
)
fs <- tempfile(fileext = ".vtr")
write_vtr(people, fs)

tbl(fs) |>
  mutate(
    name_len = nchar(name),
    short_city = substr(city, 1, 3),
    has_a = grepl("a", city)
  ) |>
  collect()
```

## Aggregation

Group data with `group_by()` and compute summaries with `summarise()`.
Supported aggregation functions: `n()`, `sum()`, `mean()`, `min()`, `max()`, `sd()`, `var()`, `first()`, `last()`, `any()`, `all()`, `median()`, `n_distinct()`.

```{r summarise}
tbl(f) |>
  group_by(cyl) |>
  summarise(
    count = n(),
    avg_mpg = mean(mpg),
    total_hp = sum(hp),
    best_mpg = max(mpg)
  ) |>
  collect()
```

For quick counts, use `count()` or `tally()`:

```{r count}
tbl(f) |>
  count(cyl, sort = TRUE) |>
  collect()
```

```{r tally}
tbl(f) |>
  group_by(gear) |>
  tally() |>
  collect()
```

Use `sd()`, `var()`, `first()`, and `last()` for more detailed summaries:

```{r agg-advanced}
tbl(f) |>
  group_by(cyl) |>
  summarise(
    mpg_sd = sd(mpg),
    mpg_var = var(mpg),
    first_hp = first(hp),
    last_hp = last(hp)
  ) |>
  collect()
```

## Sorting and slicing

`arrange()` sorts rows. Wrap column names in `desc()` for descending order:

```{r arrange}
tbl(f) |>
  select(mpg, cyl, hp) |>
  arrange(cyl, desc(mpg)) |>
  collect() |>
  head(8)
```

`slice_head()` returns the first n rows. `slice_min()` and `slice_max()` use
an optimized top-N algorithm to avoid a full sort:

```{r slice}
# First 5 rows
tbl(f) |>
  slice_head(n = 5) |>
  collect()
```

```{r slice-min}
# 3 cars with lowest mpg
tbl(f) |>
  select(mpg, cyl, hp) |>
  slice_min(order_by = mpg, n = 3) |>
  collect()
```

By default, `slice_min()` and `slice_max()` include ties. Use `with_ties = FALSE`
for exactly `n` rows:

```{r slice-ties}
tbl(f) |>
  select(mpg, cyl) |>
  slice_min(order_by = cyl, n = 3, with_ties = FALSE) |>
  collect()
```

## Joins

Join two vectra tables with `left_join()`, `inner_join()`, `right_join()`,
`full_join()`, `semi_join()`, or `anti_join()`.

```{r joins}
# Create a lookup table
cyl_info <- data.frame(
  cyl = c(4, 6, 8),
  engine_type = c("inline", "v-type", "v-type"),
  stringsAsFactors = FALSE
)
f_cyl <- tempfile(fileext = ".vtr")
write_vtr(cyl_info, f_cyl)

tbl(f) |>
  select(mpg, cyl, hp) |>
  left_join(tbl(f_cyl), by = "cyl") |>
  collect() |>
  head()
```

For joins on differently named columns, use a named vector:

```{r join-named}
ratings <- data.frame(
  cylinders = c(4, 6, 8),
  rating = c("A", "B", "C"),
  stringsAsFactors = FALSE
)
f_rat <- tempfile(fileext = ".vtr")
write_vtr(ratings, f_rat)

tbl(f) |>
  select(mpg, cyl) |>
  inner_join(tbl(f_rat), by = c("cyl" = "cylinders")) |>
  collect() |>
  head()
```

## Window functions

Window functions operate row-by-row within `mutate()`. vectra supports
`row_number()`, `rank()`, `dense_rank()`, `lag()`, `lead()`, `cumsum()`,
`cummean()`, `cummin()`, and `cummax()`.

```{r window}
tbl(f) |>
  select(mpg, cyl, hp) |>
  slice_head(n = 8) |>
  mutate(
    rn = row_number(),
    prev_mpg = lag(mpg),
    next_mpg = lead(mpg),
    running_total = cumsum(hp)
  ) |>
  collect()
```

`rank()` and `dense_rank()` take a column argument:

```{r rank}
tbl(f) |>
  select(mpg, cyl) |>
  slice_head(n = 8) |>
  mutate(mpg_rank = rank(mpg)) |>
  collect()
```

## Working with dates

Date columns roundtrip through vectra. Extract components with `year()`,
`month()`, and `day()`:

```{r dates}
dates_df <- data.frame(
  event_date = as.Date(c("2020-03-15", "2020-07-01", "2021-01-15", "2021-06-30")),
  value = c(10, 20, 30, 40)
)
fd <- tempfile(fileext = ".vtr")
write_vtr(dates_df, fd)

tbl(fd) |>
  mutate(yr = year(event_date), mo = month(event_date)) |>
  group_by(yr) |>
  summarise(total = sum(value)) |>
  collect()
```

Filter by date using `as.Date()`:

```{r date-filter}
tbl(fd) |>
  filter(event_date >= as.Date("2021-01-01")) |>
  collect()
```

## Format conversion

vectra can read and write CSV files, making it easy to convert between formats.
The entire pipeline streams batch-by-batch without loading all data into memory:

```{r csv-pipeline}
# CSV -> VTR
csv_in <- tempfile(fileext = ".csv")
write.csv(mtcars, csv_in, row.names = FALSE)

vtr_file <- tempfile(fileext = ".vtr")
tbl_csv(csv_in) |> write_vtr(vtr_file)

# VTR -> CSV (with a filter applied)
csv_out <- tempfile(fileext = ".csv")
tbl(vtr_file) |>
  filter(cyl == 6) |>
  write_csv(csv_out)

# Verify the round-trip
read.csv(csv_out) |> head()
```

## Inspecting the plan

Every vectra pipeline builds a lazy execution plan. Use `explain()` to print
the plan tree before collecting, which is helpful for debugging:

```{r explain}
tbl(f) |>
  filter(cyl > 4) |>
  select(mpg, cyl, hp) |>
  arrange(desc(mpg)) |>
  explain()
```

## Previewing data

Use `glimpse()` to see column types and a preview without collecting everything:

```{r glimpse}
tbl(f) |> glimpse()
```

## Cleanup

```{r cleanup}
unlink(c(f, fs, f_cyl, f_rat, fd, csv_in, csv_out, vtr_file))
```
