---
title: "vectra: Larger-than-Memory Tabular and Spatial Analysis in R"
date: "2026-09-28"
abstract: >
  Many analyses in ecology and other data-rich fields start from tables that
  do not fit in the memory of the computer running R, and end with a model or
  summary small enough to inspect. The \CRANpkg{vectra} package connects the
  two within R. It combines a pull-based columnar query engine written in C,
  a compressed on-disk format with per-row-group statistics, streaming spatial
  operations built on GEOS, and interfaces that pass query results to R
  functions and model-fitting routines one batch or one partition at a time.
  This paper describes how the engine executes a query and states, for each
  class of operation, what it holds in memory and what it writes to disk. It
  then measures elapsed time, peak memory, and temporary disk use against
  \CRANpkg{arrow}, \CRANpkg{duckdb}, and \CRANpkg{data.table} as workloads
  grow. A case study follows 401 million European vascular-plant occurrence
  records from GBIF through cleaning, a join to climate data, and per-region
  models of how geographical range size relates to climatic niche breadth.
draft: true
author:
  - name: Gilles Colling
    affiliation: University of Vienna
    address:
    - Department of Botany and Biodiversity Research
    - Division of BioInvasions, Global Change & Macroecology
    - Rennweg 14, 1030 Vienna, Austria
    url: https://gillescolling.com
    orcid: 0000-0003-3070-6066
    email: gilles.colling051@gmail.com
type: package
output:
  rjtools::rjournal_article:
    self_contained: yes
    toc: no
bibliography: vectra.bib
header-includes:
  - \renewcommand{\thesection}{\arabic{section}}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.align = "center", out.width = "100%")
library(ggplot2)
knitr::opts_knit$set(root.dir = normalizePath("."))
source("scripts/results.R")   # loads the measured results; defines B(), fmt()
```

# Introduction {#sec:intro}

Occurrence records, vegetation plots, telemetry fixes, and remote-sensing
extracts share a practical property: the raw table is large, the analytical
result is small, and most of the work between them is filtering, joining,
grouping, and reshaping. A European extract of vascular-plant records from
the Global Biodiversity Information Facility [GBIF, @gbif2026snapshot] holds
about 400 million rows. Loaded as an R data frame with a dozen columns it
would need more than 30 GB of memory, while the species-level table that a
macroecological model uses has a few tens of thousands of rows. The question
for the analyst is how to get from one to the other on the computer at hand.

R offers several answers, each covering part of this path. \CRANpkg{data.table}
[@datatable] makes in-memory manipulation fast and memory-frugal, but the
table has to fit in memory. \CRANpkg{arrow} [@arrow] and \CRANpkg{duckdb}
[@duckdb; @raasveldt2019duckdb] bring columnar query engines to R. Both accept
\CRANpkg{dplyr} verbs [@dplyr], through arrow's own bindings and through
\CRANpkg{dbplyr} [@dbplyr] or \CRANpkg{duckplyr} [@duckplyr]. Both read
Parquet files [@parquet] larger than memory, and DuckDB spills sorts,
aggregations, and joins to disk. Earlier packages such as \CRANpkg{ff}
[@ff], \CRANpkg{bigmemory} [@bigmemory], and \CRANpkg{disk.frame}
[@diskframe] provide file-backed vectors and chunked data frames.
\CRANpkg{biglm} [@biglm] fits linear and generalised linear models in
bounded memory from data supplied in chunks. For spatial data,
\CRANpkg{sf} [@sf] and \CRANpkg{terra} [@terra] are the standard vector and
raster tools. Both assume that the working layer, or at least the part of it
being processed, is in memory.

The gaps sit at the joins between these tools. A query engine returns its
result to R in one piece, so an analysis whose next step is an R function
(a model fit, a custom summary, a per-group computation) either collects a
result that may itself be too large, or writes glue code that pages through
the result. Spatial operations on hundreds of millions of points leave the
query engine for a spatial package that holds its input in memory. And the
memory an analysis needs is rarely stated per operation, so users find out
whether a pipeline fits by running it.

\CRANpkg{vectra} [@vectra] addresses these gaps from inside R. It is an R
extension with a query engine written in C11, reached through R's `.Call()`
interface [@rext]. It provides dplyr-style verbs over a lazy query plan, a
compressed columnar file format (`.vtr`), streaming readers for CSV, SQLite,
and GeoTIFF, geometry operations on the GEOS C API [@geos] supplied by
\CRANpkg{libgeos} [@libgeos], and interfaces that hand a query result to R
one batch (`collect_chunked()`, `chunk_feeder()`) or one on-disk partition
(`offload()`, `group_map()`) at a time. The package has been on CRAN since
2026, and \CRANpkg{taxify} [@taxify] depends on it.

This paper makes three contributions:

1. **A computational design.** It shows how lazy tabular queries, streaming
   spatial operations, and incremental or partitioned model fitting connect
   through three mechanisms: batches, disk-backed intermediate results, and
   partitions (Sections \@ref(sec:example) and \@ref(sec:design)).
2. **An account of resource use.** For each class of operation it states
   what is held in memory, what spills to disk, and which quantity of the
   data (input rows, distinct groups, build-side rows, partition size, result
   size) sets the peak (Section \@ref(sec:memory)).
3. **An evaluation and a worked analysis.** It measures these claims against
   arrow, DuckDB, and data.table on a machine with 8 GB of memory
   (Section \@ref(sec:eval)). It then carries 401 million GBIF records through
   to a fitted macroecological model and reports the scientific result
   alongside its computational cost (Section \@ref(sec:case)).

Section \@ref(sec:discussion) discusses which workloads the measurements
support using vectra for, where its semantics and resource bounds stop, and
what its storage and dependency choices imply.

# A working analysis with vectra {#sec:example}

The example below is a condensed form of the case study in
Section \@ref(sec:case). It starts from a `.vtr` file of occurrence records,
attaches a climate value to each record through a grid-cell key, reduces the
records to one row per species and cell, and fits a model per European
region. File names are shortened; the complete scripts are in the
supplementary material.

```{r example, echo = TRUE, eval = FALSE}
library(vectra)

# 1. Open the records lazily. Nothing is read yet.
rec <- tbl("records.vtr")

# 2. Describe the cleaning step. Verbs build a plan; still nothing is read.
clean <- rec |>
  filter(occurrencestatus == "PRESENT", taxonrank == "SPECIES",
         is.na(unc) | unc <= 10000, year >= 1950) |>
  mutate(cell = floor((72 - lat) * 6) * 420 + floor((lon + 25) * 6)) |>
  select(specieskey, cell, datasetkey)

# 3. Execute: the plan runs batch by batch and the result streams to disk.
write_vtr(clean, "clean.vtr")

# 4. A second pipeline: one row per species and grid cell, joined to a
#    small table of cell attributes (climate, region, area).
tbl("clean.vtr") |>
  group_by(specieskey, cell) |>
  summarise(n = n(), n_ds = n_distinct(datasetkey)) |>
  inner_join(tbl("cells.vtr"), by = "cell") |>
  write_vtr("occ_env.vtr")

# 5. Split the occupancy table on disk by region, then run arbitrary R on
#    each region's shard, which is read into memory on its own.
p <- offload(tbl("occ_env.vtr"), by = "region")
fits <- group_modify(p, function(d, region) fit_one(species_table(d)))
```

Three things about this code matter for what follows. First, the verbs do
not compute anything. `tbl()` returns a `vectra_node`, a reference to a plan
tree held in C, and every verb wraps that tree in a new node. Computation
happens only at a *terminal* operation, which pulls data through the plan:
`collect()` returns a data frame, `write_vtr()`, `write_csv()`, and
`write_sqlite()` stream to a file, `collect_chunked()` folds an R function
over the batches, `chunk_feeder()` exposes the result as a resettable
generator, and `offload()` materialises the result to disk as a replayable
node or a set of partitions. A node is consumed by the terminal operation
that runs it, so a plan that will be read twice is either rebuilt or
offloaded.

Second, `explain()` prints the optimised plan without running it, so the
memory a pipeline needs can be read before it runs. Section
\@ref(sec:design) follows one plan in detail.

Third, the example never holds the record table in R. Steps 3 and 4 write
their results to disk, and step 5 reads one region's shard at a time. The
largest object in R memory is one shard of the occupancy table, which is
orders of magnitude smaller than the record table.
<!-- TODO(case study): give the largest shard's row count once 05_models.R has run. -->

# Execution and storage design {#sec:design}

## Plans, batches, and pull-based execution

A vectra plan is a tree of nodes. Each node implements a `next_batch()`
function and a destructor, following the iterator model of query evaluation
[@graefe1994volcano]. A terminal operation calls `next_batch()` on the root.
The root calls its child, and so on down to a scan, which reads one row group
from its source and returns it as a *batch*. The batch then passes back up
the tree, transformed at each node. Batches are columnar. Each column is a
typed array (64-bit integer, double, logical, or string) with a validity
bitmap for missing values, and whole columns are processed at a time rather
than one row at a time [@boncz2005monetdb]. A filter does not copy the rows
it keeps. It attaches a *selection vector* of row indices to the batch,
which downstream nodes read through. The batch size is set by the source:
the row-group size of a `.vtr` file (131,072 rows by default), 65,536 rows
for CSV and SQLite scans, and 256 raster rows for GeoTIFF scans.

Expressions in `filter()`, `mutate()`, and `summarise()` are captured in R,
serialised into a small expression tree, and evaluated in C. This covers
arithmetic, comparison and logical operators, `%in%`, `if_else()`,
`case_when()`, `coalesce()`, date-time components, string functions
(including regular expressions and edit distances), and geometry predicates
and measures (`st_area()`, `st_intersects()`, and others) computed with GEOS
on a hex-encoded well-known-binary (WKB) geometry column. Expensive per-row
operations such as regular expressions, string distances, and geometry
predicates run in parallel with OpenMP once a batch passes a size threshold.
A function the engine does not know is an error, not a silent fallback to R.

The verbs are S3 methods on `vectra_node` for generics that vectra defines
itself. vectra is not a dplyr backend and does not import dplyr. Column
selection (`select()`, `rename()`, `relocate()`, `across()`) goes through
\CRANpkg{tidyselect} [@tidyselect], and argument capture through
\CRANpkg{rlang} [@rlang].

## The optimiser

Before execution, two rewrites run over the plan. *Predicate pushdown*
attaches a filter that sits directly above a `.vtr` scan to the scan itself.
The scan can then skip whole row groups in three ways. It can use a hash
index (a `.vtri` sidecar file) for equality and `%in%` predicates. It can
binary-search a column recorded as sorted. And it can use per-row-group
minimum, maximum, and null-count statistics, which are a form of zone map
[@moerkotte1998sma]. *Column pruning* walks the tree from the root and marks
the columns each node needs, so that a `.vtr` scan decodes only those. The
optimiser also inserts hidden projections for nested aggregate expressions
such as `mean(x * y)`.

Both rewrites are local. Pushdown applies only to a filter immediately above
a scan. A filter placed after a `mutate()`, or above a `bind_rows()` of
several files, is evaluated row by row on every batch. Column pruning stops
at a `mutate()`, because the projection it creates passes every existing
column through. Section \@ref(sec:eval) measures what these limits cost.

## Following one query

```{r plan-setup, include = FALSE}
library(vectra)
set.seed(1)
f_obs <- tempfile(fileext = ".vtr"); f_site <- tempfile(fileext = ".vtr")
obs <- data.frame(site = sample(sprintf("S%03d", 1:200), 2e5, TRUE),
                  species = sample(sprintf("sp%04d", 1:500), 2e5, TRUE),
                  year = sample(1990:2024, 2e5, TRUE), cover = runif(2e5, 0, 100),
                  observer = sample(letters, 2e5, TRUE), flag = runif(2e5))
obs <- obs[order(obs$year), ]
write_vtr(obs, f_obs, batch_size = 20000)
write_vtr(data.frame(site = sprintf("S%03d", 1:200),
                     region = sample(c("N", "S", "E", "W"), 200, TRUE)), f_site)
```

Consider a table of plot observations stored in ten row groups, sorted by
year, and a small table of site attributes. The query keeps the observations
from 2015 onward, attaches each site's region, and computes mean cover per
region and species:

```{r plan, echo = TRUE}
q <- tbl(f_obs) |>
  filter(year >= 2015) |>
  select(site, species, cover) |>
  left_join(tbl(f_site), by = "site") |>
  group_by(region, species) |>
  summarise(n = n(), mean_cover = mean(cover))
explain(q)
```

Reading the plan from the leaves upwards shows where work and memory go. The
observation scan reads 4 of its 6 columns (`year` for the filter and the
three selected columns) and carries the pushed-down predicate. Because the
file is sorted by year, the row-group statistics exclude row groups that end
before 2015 without reading them (the annotation `tdc stats` marks a scan
that has these statistics). The filter and projection stream: each
batch passes through and is released. The join's right-hand (*build*) side,
200 sites, is read into a hash table before the first observation batch
arrives. The observations are the *probe* side and stream past it. The
grouped aggregation is preceded by a sort node that the engine inserts
itself: every grouped `summarise()` sorts its input by the grouping keys and
then aggregates one group at a time. This makes the number of groups
irrelevant to memory, at the cost of sorting even when there are only a few
groups (Section \@ref(sec:memory)).

## The `.vtr` format

A `.vtr` file is a sequence of row groups. Each column of each row group is
stored as one self-describing compressed block, written by the \pkg{tdc}
codec library [@tdc] that is vendored into the package. The file ends with an
index of row groups that records each block's offset, size, and statistics
(minimum, maximum, null count), followed by a 24-byte trailer holding a
64-bit digest of the file's bytes. A scan reads the index once, then reads
only the blocks of the columns and row groups it needs. By default each block
is byte-shuffled and LZ-compressed. With `compress = "small"` the writer
trial-encodes each block with several models (delta, dictionary, sparse,
floating-point prediction) and entropy coders and keeps the smallest. String columns are dictionary-encoded, so
`collect()` creates each distinct string once rather than once per row.
Dates, date-times with their time zone, and factors with their levels round
trip through column annotations.

Stores can grow without being rewritten. `append_vtr()` adds row groups, or
columns, after the existing data and rewrites only the index and trailer, so
building a store by repeated appends costs time linear in the rows written.
`delete_vtr()` records deleted rows in a sidecar tombstone file.
`create_index()` builds a hash index with one entry per distinct key and row
group. The index is stamped with the store's row count and digest, so an
index left over from a replaced store is ignored rather than used to prune
row groups that no longer match.

## Where results go

Table \@ref(tab:sinks-tab) lists the ways a query result leaves the engine. The
distinction that matters for memory is whether the result is materialised in
R as a whole, in pieces, or not at all.

```{r sinks-tab}
sinks <- data.frame(
  Interface = c("collect(), collect(), pull()", "write_vtr(), write_csv(), write_sqlite(), write_tiff()",
                "collect_chunked(x, f, .init)", "chunk_feeder(source)",
                "offload(x)", "offload(x, by = key)", "group_map(), group_modify()"),
  `Result in R` = c("whole result", "none", "one batch + accumulator", "one batch",
                    "none (spilled to .vtr)", "none (one .vtr run set per key)",
                    "one partition at a time"),
  `Typical use` = c("small results: summaries, subsets", "large results, format conversion",
                    "one-pass reductions: counts, X'X and X'y", "iterative fits: biglm::bigglm()",
                    "replay a prepared query across passes", "localise within-group dependence",
                    "per-group models or custom code"),
  check.names = FALSE)
ktab(sinks, "Terminal operations and what each holds in R memory.", widths = c("5.2cm", "3.6cm", "5.2cm"))
```

`collect_chunked()` is a left fold: it calls `f(acc, chunk)` on each batch
in turn, so any reduction with a bounded accumulator (a count, a histogram,
sufficient statistics) runs in one pass. With a `combine` function it also
folds a partition shard by shard and merges the results. `chunk_feeder()`
follows the `data(reset)` protocol of `biglm::bigglm()`. The model can
therefore drive the iteration itself and re-read the data on each pass of
iteratively reweighted least squares [@mccullagh1989glm; @miller1992].
Because a plan is consumed when it runs, `chunk_feeder()` takes either a
function that rebuilds the plan or an offloaded node. `offload(x)` writes the
result of `x` to a temporary `.vtr` file once and returns a node that replays
it, so each model pass reads the prepared columns instead of re-running the
upstream joins. `offload(x, by = key)` writes one set of run files per key
value, per value range, or per hash bucket. `group_map()` then reads each
shard into a data frame and calls an arbitrary R function on it.

## Spatial operations

vectra has no geometry type. A geometry travels as a hex-encoded WKB string
in an ordinary column, and point data may instead be given as two coordinate
columns. The streaming spatial verbs, `spatial_join()`, `spatial_filter()`,
`spatial_clip()`, and `spatial_map()` among others, differ from the tabular
verbs in one respect: they execute when called. Each pulls its input batch
by batch, processes the batch, buffers the output in R, and writes it to
temporary `.vtr` run files once the buffer passes one eighth of the memory
budget. It then returns a lazy node over those files. For recognised
predicates on planar coordinates, `spatial_join()` and `spatial_filter()`
parse the resident right-hand layer once into GEOS prepared geometries
indexed by an STR tree. Each batch is then tested in C, in parallel, without
converting it to an \pkg{sf} object. Other predicates, and geographic
coordinates with spherical geometry enabled, fall back to calling
`sf::st_join()` on each decoded batch. When both layers are too large to hold
in memory, `spatial_join(partition = grid(cellsize))` bins both layers to a
regular grid and joins one cell at a time. Raster data enter as tables
(`tbl_tiff()` yields one row per pixel), through point extraction
(`tiff_extract_points()`), or through the tiled `.vec` raster format with
strip-wise `zonal()`, `focal()`, `terrain()`, and `warp()`. The spatial verbs
need \pkg{sf} for CRS handling and for the fallback path, so \pkg{sf} is a
suggested dependency that becomes a requirement as soon as they are used.

# Memory requirements of analytical operations {#sec:memory}

Every operation in a vectra plan belongs to one of six classes, according to
what it must retain between batches (Table \@ref(tab:memclass-tab)). All
buffering subsystems derive their limit from a single budget, `vectra_mem()`.
It is set with `options(vectra.memory = "4GB")` and defaults to half of the
physical memory.

```{r memclass-tab}
mc <- data.frame(
  Class = c("Streaming", "Bounded by output parameter", "Budgeted, spills to disk",
            "Partitioned", "Resident comparison side", "Materialised in R"),
  Operations = c("scans; filter(), select(), mutate(), slice_head(); bind_rows(); write_*(); collect_chunked(); chunk_feeder()",
                 "slice_min() / slice_max() with with_ties = FALSE",
                 "arrange(); grouped summarise() incl. median() and n_distinct(); window functions; hash joins; interval and fuzzy joins; index builds",
                 "offload(by =); group_map(); spatial_dissolve(); spatial_join(partition = grid())",
                 "spatial_join(), spatial_filter(), spatial_clip() against an sf layer; materialize() blocks",
                 "collect(), collect(), pull(); fallbacks: slice_tail(), reframe(), ties in slice_min()"),
  `Peak memory set by` = c("one batch (row-group size x columns)",
                           "k rows",
                           "vectra_mem() for the buffered state, plus one batch",
                           "one shard (the largest partition), or the routing buffer",
                           "the resident layer + one batch + output buffer (vectra_mem()/8)",
                           "the size of the result"),
  check.names = FALSE)
ktab(mc, "Classes of operation by the state they retain between batches.", widths = c("2.6cm", "7cm", "4.4cm"))
```

**Streaming operations** retain nothing between batches, so their memory is
one batch regardless of how many rows pass through. A pipeline built only
from these, and ending in a file sink or a bounded fold, runs on any input
that fits on disk.

**Operations bounded by an output parameter.** `slice_min(x, n = k,
with_ties = FALSE)` keeps a heap of the `k` best rows, so its state grows
with `k`, not with the input.

**Budgeted operations** must see all input before producing some output. In
vectra they accumulate state up to `vectra_mem()` and then spill.

- `arrange()` is an external merge sort [@knuth1998taocp3]. It accumulates
  rows in column builders until their estimated size passes the budget,
  sorts them, and writes a sorted run. At the end it merges the runs with a
  heap. The number of runs merged at once is capped (at most 64), and is
  sized from the measured row width so that the merge buffers stay within
  half the budget. More runs are merged in several passes.
- A grouped `summarise()` always goes through this sort, then aggregates one
  group at a time. Memory therefore does not depend on the number of groups,
  unlike a hash aggregation. `median()` and `n_distinct()` need every value
  of a group and spill a single large group to disk as well.
- Hash joins build the right-hand table in memory and stream the left. If
  the build side exceeds the budget, both sides are hash-partitioned into 64
  run files and joined one partition at a time, as in a Grace hash join
  [@kitsuregawa1983grace; @shapiro1986join]. A partition that is still too
  large is split again with a different hash, up to three levels. A
  partition that cannot be split, because a single key value is too frequent,
  is joined by a block-nested loop over budget-sized blocks.
- Grouped window functions sort by group and row order and process one
  group at a time, so they hold the largest group. Ungrouped windows stream
  in one pass, after a sort when the function needs an ordering (ranks,
  `lead()`). Interval joins sort both sides and sweep them, holding only the
  intervals open at the current position.

In this class the budget, not the data, sets the peak. Two qualifications
apply. The budget bounds the engine's *estimate* of its buffered state. The
sort that follows accumulation also needs working space, so peak process
memory can exceed the budget (Section \@ref(sec:eval) measures by how much).
And operations that spill trade memory for disk traffic: a sort that spills
reads and writes its input at least once more.

**Partitioned operations** make one pass that routes rows to per-key run
files, then process one partition at a time. `offload(x, by = key)` first
writes `x` to a temporary file. It then reads that file batch by batch and
routes each batch's rows into per-key buffers held as R data frames,
flushing them to run files when their combined size passes
`vectra_mem() / 8`. The routing cost is therefore paid in R. It also
produces one run file per key per flush, so a key with many values produces
many small files. `group_map()` and `group_modify()` then load one shard at a
time, and their peak is the largest shard as an R data frame.

**Resident comparison sides.** A spatial join against an \pkg{sf} layer
keeps that layer, its prepared geometries, and its spatial index in memory,
together with one input batch and the output buffer.

**Materialisation in R.** `collect()` returns the whole result as a data
frame, and it is the only terminal operation whose memory grows with the
result. It is safe for summaries and selective subsets. For results of
roughly the input's size, a file sink or a fold is the bounded alternative.
Its cost is not only the data. An R data frame stores each value in R's own
representation (8 bytes per double, a pointer per string), so a collected
result is usually larger than its compressed `.vtr` form. A few operations
fall back to R and collect their input: `slice_tail()`, `reframe()`,
`slice_min()` with `with_ties = TRUE` (the dplyr default), and `distinct()`
with `.keep_all = TRUE` on a column subset.

These classes compose. The memory of a pipeline is the sum over its
simultaneously active stateful nodes, plus one batch per streaming stage.
Because the stateful nodes share one budget knob rather than one budget, a
plan with a sort and a spilling join can hold up to twice the budget.
`explain()` marks each node as streaming or materialising, which is enough
to find the nodes that count.

```{r eval-section, child = "sections/evaluation.Rmd"}
```

```{r case-section, child = "sections/case-study.Rmd"}
```

```{r discussion-section, child = "sections/discussion.Rmd"}
```

# Computational details {#sec:details}

The results in this paper were obtained with R `r R_VERSION` [@rcore] and
vectra `r VECTRA_VERSION`, arrow `r ARROW_VERSION`, duckdb
`r DUCKDB_VERSION`, data.table `r DT_VERSION`, sf `r SF_VERSION` (GEOS
`r GEOS_VERSION`), and biglm `r BIGLM_VERSION`, on Ubuntu 24.04 in a virtual
machine with two virtual CPUs (Intel Xeon, 2.1 GHz), 8 GB of memory, no
swap, and local SSD storage. The benchmark harness, the case-study scripts,
and the raw measurements are in the supplementary material and in the
package repository at <https://github.com/gcol33/vectra>. vectra is
available from CRAN at <https://CRAN.R-project.org/package=vectra>.

# Acknowledgements

<!-- TODO(Gilles): funding, colleagues, GBIF data publishers. -->
We thank the data publishers who make their occurrence records available
through GBIF.
